---
title: Just how good was Georgia's defense in '21
author: Kirk Rangel
date: '2022-04-09'
slug: just-how-good-was-georgia-s-defense-in-21
categories: []
tags: []
---

The main topic heading into the 2021 College Football Playoff games was none other than Georgia's defense. After putting out a dominant performance in the regular season the question became: Is Georgia's defense able slow down and stop the powerful offense of Alabama. Even after a loss to the Crimson Tide (41-24) in the regular season Georgia was able to get back to its dominant self when the National Championship was on the line. With all the hype on Georgia, just how good was Georgia's defense?
```{r message=FALSE, warning=FALSE, echo=FALSE}
library(tidyverse)
library(scales)
library(ggrepel)

fb <- read_csv("footballlogs21.csv")
```

The first question we'll ask and answer is how does Georgia's differential of efficiency on both sides of the ball stacks up against others as the season progressed. Finding this differential by taking the defense average of total plays divided by average yards gained and subtracting that from the offense average of total plays divided by average yards.
```{r message=FALSE, warning=FALSE, echo=FALSE}
diffFB <-fb %>%
  mutate(Differential = OffenseAvg - DefAvg) %>%
  group_by(Team) %>%
  mutate(AvgDiff = cumsum(Differential))
```

To get a good sense of where Georgia lands we'll compare them against the rest of the AP Top 10 Ranked teams at the end of the season.
```{r message=FALSE, warning=FALSE, echo=FALSE}
apTop <- diffFB %>%
  filter(Team =="Baylor"|Team=="Ohio State" |Team=="Oklahoma State" | Team =="Notre Dame" |Team =="Michigan State"
         | Team=="Oklahoma")
```

We'll also spotlight the final four teams in the College Football Playoffs.
```{r message=FALSE, warning=FALSE, echo=FALSE}
uadiff <- diffFB %>% filter(Team=="Alabama")
ugdiff <- diffFB %>% filter(Team=="Georgia")
mudiff <- diffFB %>% filter(Team=="Michigan")
cindiff <- diffFB %>% filter(Team=="Cincinnati")
```

Here's where Georgia lands in the top 10.
```{r message=FALSE, warning=FALSE, echo=FALSE}
ggplot() + geom_step(data=apTop, aes(x=Date, y=AvgDiff, group=Team), color = "grey") +
  geom_step(data=uadiff, aes(x=Date, y=AvgDiff, group=Team), color = "#9E1B32") +
  geom_step(data=ugdiff, aes(x=Date, y=AvgDiff, group=Team), color = "#BA0C2F") +
  geom_step(data=mudiff, aes(x=Date, y=AvgDiff, group=Team), color = "#00274C") +
  geom_step(data=cindiff, aes(x=Date, y=AvgDiff, group=Team), color = "#E00122") +
    annotate("text", x=(as.Date("2021-12-20")), y=43, label="Georgia") +
  annotate("text", x=(as.Date("2021-12-20")), y=37, label="Cincinnati") +
  annotate("text", x=(as.Date("2021-12-20")), y=29, label="Alabama") +
  annotate("text", x=(as.Date("2021-12-20")), y=22, label="Michigan") +
  labs(
    x="Date", 
    y="Cumulative Point Differential", 
    title="Georgia's Rise To The Top On Both Sides Of The Ball", 
    subtitle="Georgia's defense getting its team differential ahead in the AP Top 10", 
    caption="Source: http://mattwaite.github.io/sportsdatafiles/footballlogs21.csv | By Kirk Rangel") +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 18, face = "bold"),
    axis.title = element_text(size = 10), 
    plot.subtitle = element_text(size=10), 
    panel.grid.minor = element_blank()
    ) 
```

Georgia quickly got out and ahead early in the season and climbed with momentum. Their defense was able to keep their opponents to a minimal amount of plays and yards while getting their offense on the field more.

Next we'll take a more in depth look at just how efficient Georgia's defense was. Where do they rank when taking into account all aspects of defense? And as they saying goes, does defense win games?

We'll total up each teams win from the season.
```{r message=FALSE, warning=FALSE, echo=FALSE}
winlossfb <- fb %>%
  mutate(
    wins = case_when(
      grepl("W", Outcome) ~1,
      grepl("L", Outcome) ~0))
```

As well as adding up team average totals including pass and rush percentages, opponent first downs, opponent turnovers, opponent total score, pass and rush touchdowns, opponent yards, and opponent plays.

We'll also do this for all things offense, we'll use it later but for now we're focused on defense.
```{r message=FALSE, warning=FALSE, echo=FALSE}
teamtotal <- winlossfb %>%
  group_by(Conference, Team) %>%
  summarise(
    PassingPctAvg = mean(PassingPct),
    PassingYdsAvg = mean(PassingYds),
    RushingYdsAvg = mean(RushingYds),
    OffPlaysAvg = mean(OffensivePlays),
    OffYdsAvg = mean(OffensiveYards),
    FirstDownAvg = mean(FirstDownTotal),
    TurnoverAvg = mean(TotalTurnovers),
    ScoreAvg = mean(TeamScore),
    TotalScore = sum(TeamScore),
    TotalOppScore = sum(OpponentScore),
    OppAvgScore = mean(OpponentScore),
    OppPassPctAvg = mean(DefPassingPct),
    OppPassYdsAvg = mean(DefPassingYds),
    OppPassTDAvg = mean(DefPassingTD),
    OppRushYdsAvg = mean(DefRushingYds),
    OppRushTDAvg = mean(DefRushingTD),
    OppPlaysAvg = mean(DefPlays),
    OppYdsAvg = mean(DefYards),
    OppFirstDownAvg = mean(DefFirstDownTotal),
    OppTurnoversAvg = mean(DefTotalTurnovers),
    TotalWins = sum(wins))
```

To get an efficiency rating we'll create a z-score, a statistical measure of the distance from the mean called a z-score, taking all the defensive categories and adding them together. The lower the score for each category the better the total rating will be for a team. Also doing this for the offensive side of things. 
```{r message=FALSE, warning=FALSE, echo=FALSE}
teamzscore <- teamtotal %>%
  mutate(
    PTSallowedZScore = as.numeric(scale(OppAvgScore, center = TRUE, scale = TRUE))* -1,
    OppPassCmpPctZScore = as.numeric(scale(OppPassPctAvg, center = TRUE, scale = TRUE))* -1,
    OppPassYdsZScore = as.numeric(scale(OppPassYdsAvg, center = TRUE, scale = TRUE))* -1,
    OppRushYdsZScore = as.numeric(scale(OppRushYdsAvg, center = TRUE, scale = TRUE))* -1,
    OppPlaysZScore = as.numeric(scale(OppPlaysAvg, center = TRUE, scale = TRUE))* -1,
    OppTotalYdsZScore = as.numeric(scale(OppYdsAvg, center = TRUE, scale = TRUE))* -1,
    OppFirstDownZScore = as.numeric(scale(OppFirstDownAvg, center = TRUE, scale = TRUE))* -1,
    OppTurnoversZScore = as.numeric(scale(OppTurnoversAvg, center = TRUE, scale = TRUE))* -1,
    DefensiveEfficiency= PTSallowedZScore + OppPassCmpPctZScore + OppPassYdsZScore + OppRushYdsZScore + OppPlaysZScore + OppTotalYdsZScore + OppFirstDownZScore + OppTurnoversZScore,
    PassPctZScore = as.numeric(scale(PassingPctAvg, center = TRUE, scale = TRUE)),
    PassYdsZScore = as.numeric(scale(PassingYdsAvg, center = TRUE, scale = TRUE)),
    RushYdsZScore = as.numeric(scale(RushingYdsAvg, center = TRUE, scale = TRUE)),
    OffPlayZScore = as.numeric(scale(OffPlaysAvg, center = TRUE, scale = TRUE)),
    OffYdsZScore = as.numeric(scale(OffYdsAvg, center = TRUE, scale = TRUE)),
    FirstDownZScore = as.numeric(scale(FirstDownAvg, center = TRUE, scale = TRUE)),
    TurnoverZScore = as.numeric(scale(TurnoverAvg, center = TRUE, scale = TRUE)),
    PTSZScore = as.numeric(scale(ScoreAvg, center = TRUE, scale = TRUE)),
    OffensiveEfficiency= PassPctZScore + PassYdsZScore + RushYdsZScore + OffPlayZScore + OffYdsZScore + FirstDownZScore + TurnoverZScore +PTSZScore
  )
```

To broaden the range of teams and really get a feeling of how good Georgia was we'll look at the top 10 teams from each conference.
```{r message=FALSE, warning=FALSE, echo=FALSE}
top <- teamzscore %>% 
  arrange(desc(DefensiveEfficiency)) %>%
  top_n(10)
```

Again we'll highlight the College Football Playoff teams to see where they ranked.
```{r message=FALSE, warning=FALSE, echo=FALSE}
ug <- top %>% filter(Team == "Georgia")
ua <- top %>% filter(Team == "Alabama")
mu <- top %>% filter(Team =="Michigan")
cin <- top %>% filter(Team =="Cincinnati")
```

Here's where Georgia's defensive efficiency falls among teams from each conference.
```{r message=FALSE, warning=FALSE, echo=FALSE}
ggplot() + geom_point(data=top, aes(x=DefensiveEfficiency, y =TotalOppScore, size=TotalWins),
                      color="grey",
                      alpha = .5) +
  scale_size(range = c(1, 8), name="Wins") +
  geom_point(
    data=ug,
    aes(x=DefensiveEfficiency, y =TotalOppScore, size=TotalWins),
    color = "#BA0C2F"
  ) +
  geom_point(
    data=cin,
    aes(x=DefensiveEfficiency, y =TotalOppScore, size=TotalWins),
    color = "#E00122"
  ) +
  geom_point(
    data=ua,
    aes(x=DefensiveEfficiency, y =TotalOppScore, size=TotalWins),
    color = "#9E1B32"
  ) +
  geom_point(
    data=mu,
    aes(x=DefensiveEfficiency, y =TotalOppScore, size=TotalWins),
    color = "#00274C"
  ) +
  geom_vline(xintercept = 0.6016113) +
  geom_hline(yintercept = 341.6075)+
  geom_text(aes(x=6, y=170, label="Good Defense, Less Points Allowed"), color="blue") +
  geom_text(aes(x=-5, y=500, label="Bad Defense, More Points Allowed"), color="blue") +
  annotate("text", x=9.6, y= 135, label="Georgia") +
  annotate("text", x=6.9, y= 245, label="Michigan") +
  annotate("text", x=2.8, y= 213, label="Cincinnati") +
  annotate("text", x=2.8, y= 290, label="Alabama") +
  labs(
    x="Defensive Efficiency Score ",
    y="Total Points Allowed",
    title="Solid Defense Leads to Championships",
    subtitle="'Defense wins games' looks promising with Georgia's dominanting performance",
    caption="Source: http://mattwaite.github.io/sportsdatafiles/footballlogs21.csv | By Kirk Rangel"
  ) +
  theme_light()+
  theme(
    plot.title = element_text(size = 20, face = "bold"),
      axis.title = element_text(size = 8), 
    plot.subtitle = element_text(size=10),
    panel.grid.minor = element_blank())
```

Georgia once again separates itself from everyone else coming in with the highest defensive efficiency rating while allowing the fewest points and a high win column. The other three College Football Playoff teams, all clumped together.

Even with an outstanding defense a team still needs offense to get points on the board. Finally, we'll look at the top 4 teams composite defensive ratings compared to their offensive ratings from earlier. 
```{r message=FALSE, warning=FALSE, echo=FALSE}
offdef <- top %>%
  filter(Team == "Georgia"| Team == "Alabama"| Team == "Cincinnati"| Team == "Michigan")%>%
  select(Conference, Team,DefensiveEfficiency, OffensiveEfficiency)
```

```{r message=FALSE, warning=FALSE, echo=FALSE}
zscorewide <- offdef %>%
  pivot_longer(
    cols=ends_with("Efficiency"),
    names_to="Type",
    values_to="ZScore"
  )
```

Here's another look at Georgia's composite defensive efficiency rating and how the top 4 stack up against each with their offensive power.
```{r message=FALSE, warning=FALSE, echo=FALSE}
ggplot() + 
  geom_bar(data=zscorewide, aes(x=reorder(Team, ZScore), weight=ZScore, fill=Type)) +
  coord_flip() +
  scale_fill_manual(values = c("#D2042D", "#3D7B36"))+
  labs(
    x="Team",
    y="Efficiency Score",
    title="Georgia Defense Makes Up For Not So Strong Offense",
    subtitle="Extremely efficient Georgia defense gets their offense on the field more",
    caption ="Source: http://mattwaite.github.io/sportsdatafiles/footballlogs21.csv | By Kirk Rangel"
  ) +
 theme_light()+
  theme(
    plot.title = element_text(size = 12, face = "bold"),
      axis.title = element_text(size = 8), 
    plot.subtitle = element_text(size=9),
    panel.grid.minor = element_blank())
```

Even though Georgia didn't have the best offensive score their dominant defense was able to overcome Alabama's huge offense and win a national championship. Georgia defense was able to create more opportunities for its offense, keeping them on the field more often, converting to points and wins. 

Just how good was Georgia's defense? Dominant and effective.

NOTE: This is the companion notebook for http://kirkrangel.github.io/2022/04/09/just-how-good-was-georgia-s-defense-in-21/


